---
title: "iMath (mathematical operations inside ANTsR)"
author: "Brian B. Avants, Arman Eshaghi, Jeffrey T. Duda"
date: "`r Sys.Date()`"
bibliography: REFERENCES.bib
output: rmarkdown::html_vignette
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{iMath}
    \usepackage[utf8]{inputenc}
---


```{r global options, include=FALSE}
library( knitr )
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(ANTsR)
```

## Basic operations

`iMath`  wraps some functions of `ImageMath` from [ANTs software](http://stnava.github.io/ANTs/). Basic
arithmetics (e.g., add, subtract), are built into the `antsImage` class capabilities, and are similar to
array operations in R:

```{r basic}
    fi  <- antsImageRead(getANTsRData("r16"), 2)
    sumfi <- fi + fi
    mulfi <- fi * 2
    #logarithm of image where non-zero
    logfi <- log(fi[fi>0])
    expfi <- exp(fi[fi>0])
    divfi <- sumfi / mulfi
```

##Morphological operations on masks and images

The typical rule for dilation and erosion of images in ANTsR is minimum
for erosion and maximum for dilation (ITK rules).

  * Grayscale dilation of an image (compare it with binary dilation below):

    ```{r greydilate}
    #dilating by a radius of 3 voxels
    GDdilated <- iMath(fi,"GD", 3)
    #to see what dilation has done
    invisible(plot(GDdilated))
    #to see the difference a dilation can make
    invisible(plot(GDdilated - fi))
    ```

  * Grayscale erosion of an image (compare it with binary erosion below).

    ```{r greyerode}
    erosion <- iMath(fi,"GE", 3)
    invisible(plot(erosion))
    ```

  * Morphological dilation works on binary images, where it makes white
  regions bigger:

    ```{r dilate}
    mask <- getMask(fi)
    MD <- iMath(mask, "MD", 3)
    #to see the difference it made
    invisible(plot(MD - mask))
    ```

  * Morphological erosion makes white regions of a binary mask smaller.

    ```{r erode}
    ME <- iMath(mask, "ME", 3)
    invisible(plot(ME))
    ```

  * Morphological closing of a binary image: operation `MC` fill holes with the
    provided radius parameter, for example:

    ```{r closer}
    newMask <- iMath(mask, "MC", 4)
    invisible(plot(newMask,slices=c(1,1)))
    ```

  * Morphological opening of a binary image: removes small structures at the boundary
    or interior of an image. Syntax is similar to the previous procedure with the operation
    `MO`.


##Padding or cropping an image

`PadImage` is a function to add/remove voxels to/from the boundaries of an image.

* Positive numbers will pad image in each direction. So for example using 2
  will increase an image with 256 voxels to 260.

    ```{r pad_up}
    padded <- iMath(fi, "PadImage", 2)
    #compare padded image dimensions with the original dimensions
    dim(fi)
    dim(padded)
    ```

  * Negative numbers will crop the image in each direciton. For example using -2
  will convert a 2-dimensional image with 256 voxels in X and Y direction to
  252 voxels in each dimension.

    ```{r pad_down}
    cropped <- iMath(fi, "PadImage", -2)
    #compare cropped image with the original one
    dim(fi)
    dim(cropped)
    ```

##Distance map
Distance maps may be used for a number of purposes, including: shape
analysis, skeletonization and path finding.

  * `MaurerDistance` implements ITK's [SignedMaurerDistanceMap](http://www.itk.org/Doxygen/html/classitk_1_1SignedMaurerDistanceMapImageFilter.html) which
  calculates the Euclidean distance transform of a binary
  image in time linear with total number of voxels [@Maurer2003].
  It assumes that inside the binary image has negative distance,
  while outside of the binary mask has positive distance.

    ```{r maurer}
    distanceMap <- iMath(mask, "MaurerDistance")
    invisible(plot(distanceMap))
    ```

 *  `D` implements ITK's [DanielssonDistanceMap](http://www.itk.org/Doxygen/html/classitk_1_1DanielssonDistanceMapImageFilter.html). It calculates the Euclidean
   distance map, which shows at each voxel the shortest distance to the nearest
   voxel in the background (assuming that the input is a binary image)
   [@Danielsson1980].

    ```{r distancemap}
    distanceMap <- iMath(mask, "D")
    invisible(plot(distanceMap))
    ```

##Denoising with anisotropic diffusion

To reduce noise without changing important parts of an image in computer vision, Perona-Malik is a
promising method. Perona-Malik method to reduce noise with anisotropic diffusion is accessible
via `PeronaMalik` operation in `iMath`. It requires two parameters: 1) number of iterations, and
2) conductance. This implements ITK's [GradientAnisotropicDiffusionImageFilter](http://www.itk.org/Doxygen/html/classitk_1_1GradientAnisotropicDiffusionImageFilter.html).

The conductance parameter is described by ITK's [AnisotropicDiffusionFunction](http://www.itk.org/Doxygen/html/classitk_1_1AnisotropicDiffusionFunction.html): "The conductance parameter controls the sensitivity of the conductance term in the basic anisotropic diffusion equation. It affects the conductance term in different ways depending on the particular variation on the basic equation. As a general rule, the lower the value, the more strongly the diffusion equation preserves image features (such as high gradients or curvature). A high value for conductance will cause the filter to diffuse image features more readily. Typical values range from 0.5 to 2.0 for data like the Visible Human color data, but the correct value for your application is wholly dependent on the results you want from a specific data set and the number or iterations you perform."

```{r pm}
denoised <- iMath(fi, "PeronaMalik", 10, 0.5)
invisible(plot(denoised))
# to see what the filter has removed
invisible(plot(fi - denoised))
```

## Magnitude of gradient computation

`Grad` implements ITK's [GradientMagnitudeRecursiveGaussian](http://www.itk.org/Doxygen/html/classitk_1_1GradientMagnitudeRecursiveGaussianImageFilter.html#a170eaa3168a017c92cc0b640943e9ca2) which calculates the gradient of the magnitude of an image by convolution
with the first derivative of a Gaussian. Parameters are:

* `sigma` (Optional: double, default=0.5) is the full width at half max of the Gaussian kernel,
specified in physical space units.

* `normalize` (Optional: 0 or 1 boolean, default=0) specifies if the output should be scaled to lie in [0,1]

```{r grad}
  grad <- iMath(fi, "Grad", 1)
  invisible(plot(grad))
```

## Laplacian of Gaussian of an image

`Laplacian` implements ITK's [LaplacianRecursiveGaussianImageFilter](http://www.itk.org/Doxygen/html/classitk_1_1LaplacianRecursiveGaussianImageFilter.html) which  calculates the the Laplacian of Gaussian of an image by convolving with
the second derivative of a Gaussian. Parameters are:

* `sigma` (Optional: double, default=0.5) is the full width at half max of the Gaussian kernel,
specified in physical space units.

* `normalize` (Optional: 0 or 1 boolean, default=0) specifies if the output should be scaled to lie in [0,1]

```{r lap}
  laplacianImage <- iMath(fi, "Laplacian", 1, 1)
  invisible(plot(laplacianImage))
```

##Sequential operations on images

Usually it is easier to perform sequential procedures starting from left to right,
instead of right to left, as one needs with functions. This has been made possible by
another package that `ANTsR` depends on, `magrittr`. For example, instead of:

```{r}
fi<-antsImageRead( getANTsRData("r16") , 2 )
result <- iMath(iMath(fi, "Laplacian", 1), "GD", 3)
```

  One can do:

```{r}
require(magrittr)
result <- fi %>% iMath("Laplacian",1)  %>% iMath("GD",3)
```

##Other operations

| Operation     | Example                   | Description                |
| ------------- |:-------------------------:| --------------------------:|
| FillHoles     |`img %>% iMath("FillHoles")`| Fills holes in binary object|
| GetLargestComponent |`img %>% iMath("GetLargestComponent")`|Returns largest portion of binary object|
|Normalize| `img %>% iMath("Normalize")` |Creates image negative|
|TruncateImageIntensity| `img %>% iMath("TruncateImageIntensity", 0.05, 0.95)`|Trims intensities by quantiles|
|Sharpen| `img %>% iMath("Sharpen")` | Makes edges sharper|



## All iMath operations
```{r, results='asis',echo=FALSE}
data("iMathOps")
kable( iMathOps , caption = "Valid iMath Operations", padding = 0 )
```

## References
