---
title: "Multichannel image proccessing"
author: "Jeffrey T. Duda"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{Multichannel images}
    \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE, include = FALSE}
library( knitr )
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(ANTsR)
library(ggplot2)
library(grid)

if("parallel" %in% rownames(installed.packages()) == TRUE) {
  library(parallel)
}  

```

## Overview
This document provides some examples illustrating how [ANTsR](https://github.com/stnava/ANTsR)
may be used to work with multi channel images, such as rgb (i.e. color) data. This
is still an extremely new feature and much of ANTsR does not yet support this image type.
Examples will be added to this document as new functionality is implemented.

```{r plotColor,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=FALSE}
# How to plot a 2D color image
plotColor <- function(imgList, scale=TRUE, vectors=NULL, points=NULL, paths=NULL) {

  if (class(imgList) == "antsImage") {
    imgList = list(imgList, imgList, imgList)
  }

  direction = antsGetDirection( imgList[[1]] )

  # max in all images
  maxi = 1.0
  if ( scale )
    {
    maxi = max( unlist( lapply( imgList, function(x) { max(x) } ) ) )
    }

  rgbList = lapply( imgList, function(x) { apply(t(as.matrix(x)),2,rev) / maxi })
  rgbList = lapply( imgList, function(x) { t(as.matrix(x)) / maxi })

  col <- rgb(rgbList[[1]], rgbList[[2]], rgbList[[3]])

  d = dim(rgbList[[1]])

  x = rep(1:d[2],each=d[1])
  y = rep(1:d[1], d[2])
  pts = antsTransformIndexToPhysicalPoint( imgList[[1]], cbind(x,y) )

  dat = data.frame(x=pts[,1], y=pts[,2], col=col)
  x1 = min(pts[,1])
  x2 = max(pts[,1])
  y1 = min(pts[,2])
  y2 = max(pts[,2])

  g = ggplot(dat) + geom_raster(aes(x=x, y=y, fill=col), hjust=0, vjust=0, alpha=1) + theme(legend.position="none", aspect.ratio=1,text=element_blank(),axis.ticks=element_blank(), panel.grid=element_blank() ) + scale_fill_manual(values=as.character(levels(factor(col))) )

  g = g + coord_cartesian( xlim=c(x1,x2), ylim=c(y1,y2) )
  if ( direction[1,1] > 0 ) {
    g = g + scale_x_continuous( lim=c(x1,x2) )
    }
  else {
    g = g + scale_x_reverse( lim=c(x2,x1) )  
   }
  if ( direction[2,2] > 0 ) {
    g = g + scale_y_continuous( lim=c(y1,y2) )
    }
  else {
    g = g + scale_y_reverse( lim=c(y2,y1) )  
   }

  if ( !is.null(points) ) {
    pdat = data.frame( x=points[,1], y=points[,2], id=factor(1:dim(points)[1]) )
    g = g + geom_point( data=pdat, aes(x=x, y=y, colour=id ))
  }

  if ( !is.null(paths) ) {
    g = g + geom_path(data=paths, aes(x=x,y=y,group=id,colour=id))
    }

  if ( !is.null(vectors) ) {
    xvec = as.vector( t(as.matrix(vectors[[1]])) )
    yvec = as.vector( -t(as.matrix(vectors[[2]])) )
    vpts = antsTransformIndexToPhysicalPoint( imgList[[1]], cbind(x+0.5,y+0.5) )

    mag = sqrt(xvec*xvec + yvec*yvec)
    elim = which(mag < 0.01)
    if (length(elim) > 0 ) {
      xvec = xvec[-elim]
      yvec = yvec[-elim]
      vpts = vpts[-elim,]
      }
    vdat = data.frame(x=vpts[,1]-xvec, y=vpts[,2]-yvec, xend=vpts[,1]+xvec, yend=vpts[,2]+yvec)
    g = g + geom_segment(data=vdat, aes(x=x,y=y,xend=xend,yend=yend), colour="white", alpha=0.5)
  }

  suppressWarnings(print(g))
}
```

## Basics
Because ANTsR relies upon [ITK](http://www.itk.org) for image IO, multichannel support is inherently built in.
Basic conversions and math operations work as expected.

```{r io,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
# Read in and display header info
img = antsImageRead( getANTsRData("decslice"))
img
img[64,64]

# Convert to an array
arr = as.array(img)
dim(arr)

# Convert array to multichannel antsImage
img2 = as.antsImage(arr, components=TRUE)
img2

# Convert array to antsImage and set header info via reference image
img2 = as.antsImage(arr, components=TRUE, reference=img )
img2

# Basic math ops work the same as for scalar images
2 * sum(img)
sum(img * 2)
```

Many functions only work on scalar images, so here we split a multichannel image into a list
of scalar images, then the `lapply` function provides a convenient way to process all channels.

```{r lapply,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
# Convert to list of scalar images
iList = splitChannels(img)
plotColor(iList)

sList = lapply( iList, function(x) { smoothImage(x, 1.5) } )
plotColor(sList)

# Merge back into multichannel image
simg = mergeChannels(sList)

# Write to file
# antsImageWrite(simg, "smoothslice.nii.gz")
```

It is worth mentioning that ANTsR is not yet compatible with `mclapply` as provided
in the "parallel" package. It's something we will be looking into in the future. The
main problem has to do with allocating new images, some functions that only access
existing data will work, however it does not guarantee a faster execution time.
```{r mclapply,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
iMeans = 0
time1 = 0
timeParallel = NA
if("parallel" %in% rownames(installed.packages()) == TRUE) {
  ptm = proc.time()
  iMeans = mclapply( iList, function(x) mean(x), mc.cores=min(detectCores(),2) )
  timeParallel = proc.time() - ptm
  unlist(iMeans)
}

ptm = proc.time()
iMeans = lapply(iList, function(x) mean(x) )
timeSerial = proc.time() - ptm
unlist(iMeans)

timeParallel
timeSerial
```

## Working with diffusion tensor images
```{r dti1,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
dt = antsImageRead(getANTsRData("dtislice"))
dtList = splitChannels(dt)

# Component of tensor are stored as: XX, XY, XZ, YY, YZ, ZZ
trace = dtList[[1]] + dtList[[4]] + dtList[[6]]
trace[trace<0] = 0
plotColor( trace )
```

We only want to deal with voxels in the brain, so the `getMask` function is used
to obtain rough estimate of the brain, and the image is plotted with background voxels
tinted red.

```{r dti2,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
mask = getMask(trace)
plotColor( list(trace, trace*mask, trace*mask))
```

Since we are only interested in voxels in the brain, a matrix
is created where each row is voxel in the brain and each column in a tensor component,
listed in upper.tri order. This allows us to quickly calculate the eigen
decomposition for each tensor.

```{r dti3,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
# list-based call to convert values to an array
mat = do.call(rbind, lapply( dtList, function(x) { x[mask>0] } ) )

# simplest convesion to array
# mat = as.array(dt, mask == 1)

# convert tensor from vector to matrix
initTensor <- function(x) {
  tens = diag(3)
  tens[lower.tri(tens, diag=TRUE)] = x
  tens = tens + t(tens) - diag(diag(tens))
  return(tens)
  }

getFractionalAnisotropy <- function(evs) {
  numer = sqrt( (evs[1]-evs[2])^2 + (evs[2]-evs[3])^2 + (evs[3]-evs[1])^2 )
  denom = sqrt( sum(evs*evs) )
  fa = 0
  if ( denom > 0 ) {
    fa = sqrt(0.5) * numer/denom
  }
  fa
  }

# Eigen decomposition for each tensor
# eigs = unlist( apply(mat, 2, function(x) {
#   eigen( initTensor(x) )
#   } )
#   )

all_eigs = apply(mat, 2, function(x) {
  eigen( initTensor(x) )
  } )

eigs = lapply(all_eigs, function(x) {
  x$values
})
eigs = lapply(eigs, unlist)
evalMat = do.call(rbind, eigs)
evalMat[evalMat < 0 ] = 0

# Create images from the eigenvalues
eval1 = makeImage(mask, evalMat[,1])
eval2 = makeImage(mask, evalMat[,2])
eval3 = makeImage(mask, evalMat[,3])

plotColor( list(eval1, eval2, eval3))
```

The Fractional Anisotropy (FA) is typically used to measure how directionally specific
the diffusion of water is within a voxel. This values is plotted below

```{r dti4,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
faValues = apply(evalMat, 1, getFractionalAnisotropy )
faValues[faValues < 0] = 0
fa = makeImage(mask, faValues)
plotColor( fa )
```

Also of great interest is the eigenvector associated with the largest eigenvalue, this
estimates the primary direction of diffusion (PDD) and in white matter this corresponds to
the direction that is parallel to the myelinated axons in a fiber bundle

```{r dti5,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
eig_vecs = lapply(all_eigs, function(x) {
  x$vectors
})
eig_vecs = lapply(eig_vecs, c)
vecMat = do.call(rbind, eig_vecs)


evecx = makeImage(mask, vecMat[,1])
evecy = makeImage(mask, vecMat[,2])
evecz = makeImage(mask, vecMat[,3])
decList = list(evecx, evecy, evecz)
decList = lapply(decList, abs)

plotColor( decList )
```

Weighting the magnitude of the PDD by the FA is standard practice as it highlights
regions with high directional specificity (i.e. white matter).
This is known as a directional encoded colormap (DEC) [@Pajevic1999]

```{r dti6,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
decList = lapply( decList, function(x){ fa*x } )
plotColor(decList)
```

To verify our processing, it can be useful to plot line segments showing the direction
of some of the vectors. This is done in a subset of the image to avoid too much visual
clutter.

```{r dti7,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
lower = c(40,80)
upper = c(80,120)
subfa = cropIndices(fa, lowerind = lower, upperind = upper)
subVecList = list(evecx, evecy, evecz)
subVecList = lapply( subVecList, function(x){ cropIndices(x, lowerind = lower, upperind = upper)})
subDecList = lapply( subVecList, function(x){ abs(subfa*x) })

plotColor( subDecList, vectors=list(subVecList[[1]], subVecList[[2]]) )
```

## Deterministic Fiber tractography

A common use for DTI is fiber tractography [@Basser2000]. Here we give a simplified example in which
we restrict the tracts to lie within the slice. The first step is define a set of seeds
which serve as the starting points for our tractography. Often, all points in the white
matter are used as seed for "whole-brain tracking." Here we manually define a small set
of seed points for illustration.

```{r seeds,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}

# indices of seed points
seedIndices = rbind( c(61,92), c(62,92), c(63,92) )
seedIndices = rbind( seedIndices, c(64,92), c(65,92), c(65,92) )
seedIndices = rbind( seedIndices, c(61,93), c(62,93), c(63,93) )
seedIndices = rbind( seedIndices, c(64,93), c(65,93), c(65,93) )

# convert to physical space points
seedPts = antsTransformIndexToPhysicalPoint(fa, seedIndices)
plotColor( subfa, points=seedPts )
```


```{r tracking,message=FALSE,warnings=FALSE, fig.width=7, fig.height=5, echo=TRUE}
trackFromSeed <- function(vecs, fa, seed) {
  stepSize = 0.2
  faThresh = 0.2
  iPt = seed
  pts = iPt
  idx =  antsTransformPhysicalPointToIndex(fa, seed)
  faValue = fa[idx[1], idx[2]]
  ppdx = vecs[[1]][ idx[1], idx[2] ]
  ppdy = vecs[[2]][ idx[1], idx[2] ]
  lastVec = as.vector(antsGetDirection(fa) %*% as.matrix(c(ppdx,ppdy)))
  iVec = lastVec

  while ( faValue > faThresh ) {
    iVec = iVec / sqrt(sum(iVec*iVec))
    if ( sum(iVec*lastVec) <  0 ) { iVec = -iVec }

    iPt = iPt + stepSize*iVec
    pts = rbind(pts, iPt)
    idx =  antsTransformPhysicalPointToIndex(fa, iPt)
    faValue = fa[idx[1], idx[2]]
    lastVec = iVec

    ppdx = vecs[[1]][ idx[1], idx[2] ]
    ppdy = vecs[[2]][ idx[1], idx[2] ]
    iVec = as.vector(antsGetDirection(fa) %*% as.matrix(c(ppdx,ppdy)))
  }

  iPt = seed
  idx =  antsTransformPhysicalPointToIndex(fa, seed)
  faValue = fa[idx[1], idx[2]]
  vecs = lapply( vecs, function(x){ -x })
  ppdx = vecs[[1]][ idx[1], idx[2] ]
  ppdy = vecs[[2]][ idx[1], idx[2] ]
  lastVec = as.vector(antsGetDirection(fa) %*% as.matrix(c(ppdx,ppdy)))
  iVec = lastVec

  while ( faValue > faThresh ) {
    iVec = iVec / sqrt(sum(iVec*iVec))
    if ( sum(iVec*lastVec) <  0 ) { iVec = -iVec }

    iPt = iPt + stepSize*iVec
    pts = rbind(iPt, pts)
    idx =  antsTransformPhysicalPointToIndex(fa, iPt)
    faValue = fa[idx[1], idx[2]]
    lastVec = iVec

    ppdx = vecs[[1]][ idx[1], idx[2] ]
    ppdy = vecs[[2]][ idx[1], idx[2] ]
    iVec = as.vector(antsGetDirection(fa) %*% as.matrix(c(ppdx,ppdy)))
  }

  return(data.frame(x=pts[,1],y=pts[,2]))

}

trackx = c()
tracky = c()
trackid = c()
for ( i in 1:dim(seedPts)[1] ) {
  track = trackFromSeed(list(evecx,evecy), fa, seedPts[i,])
  trackx = c(trackx,track$x)
  tracky = c(tracky,track$y)
  trackid = c(trackid, rep(i, length(track$x)))  
}
ldat = data.frame(x=trackx, y=tracky, id=factor(trackid))
plotColor(subDecList, paths=ldat, points=seedPts, vectors=list(subVecList[[1]], subVecList[[2]]))
plotColor(subfa, paths=ldat, points=seedPts)

```



## References
